Generalized Linear Mixed Models via Monte Carlo Likelihood Approximation

### 2014-03-21 GoogleProposal.tex
This fucking git thing is so stupid.
Edited my google proposal more and git is pissing me off.

Generalized linear mixed models (GLMMs) are popular in many fields from ecology to economics. Researchers clearly want to use GLMMs: a Google search for ``GLMM'' returns about 242,000 results.  The challenge for researchers is finding an easy-to-implement and reliable method for fitting and testing GLMMs. For very simple problems with just a few random effects, the likelihood can be approximated by numerical integration.

Most models have crossed random effects, which numerical integration cannot handle. Thus, a commonly used method is penalized quasi-likelihood (PQL),which is  implemented in packages such as lme4, nlme, and MASS. However, PQL relies on approximations of unknown accuracy to approximate the likelihood and suffers from problematic inferential properties, such as parameter estimates that tend to be too low (\ncite{mccu:sear:2001}). Since the likelihood is approximated to an unknown accuracy by the quasi-likelihood, any inference performed on the approximated likelihood will also produce results with an unknown level of accuracy.  Without bootstrapping, the user cannot know how valid their confidence intervals or likelihood ratio test results are when using PQL.  The popularity of PQL despite its inadequacies shows that there is a high demand for tools to fit GLMMs.

@@

My goals are  (1) to rewrite sections of my package in C to further improve its speed, (2) write functions to perform likelihood ratio tests to compare nested models, (3) write additional functions to fit models with correlated random effects.  My goals are extensions of my current  work on the topic of using MCLA to fit GLMMs.  Because I have been working on this project and am familiar with the limitations and areas of possible improvement, I am the best individual for this project.  Also, I have access to a wealth of knowledge and experience: one of my advisors is the inventor of MCLA and both of my advisors have submitted and maintained R packages.

@@

(2)  Hypothesis testing for nested models can be split into three cases: the nested models differ in their fixed effects but have the same variance components, the nested models differ by one variance component (and possibly some fixed effects), the nested models differ by two or more variance components (and possibly by some fixed effects). I have worked out the details for calculating the test statistics and p-values for the first two cases and will  be able to write the code for these two in about three weeks.  Coding the last case will take longer because I will need to determine the test statistic and its sampling distribution.

@@

In conclusion, researchers from all fields have created a high demand for a tool to fit and test GLMMs. I would like to make the process of fitting a GLMM as user-friendly and theoretically-grounded as fitting a linear model. My package will allow researchers to find maximum likelihood estimates for GLMMs with various covariance structures and perform likelihood ratio tests to compare the fit of nested models.

>>>>

Generalized linear mixed models (GLMMs) are popular in many fields from ecology to economics. The 242,000 results from a Google search for ``GLMM'' show the popularity of GLMMs.  The challenge for researchers is finding an easy-to-implement and reliable method for fitting and testing GLMMs. For very simple problems with just a few random effects, the likelihood can be approximated by numerical integration.  Most models have crossed random effects, which numerical integration cannot handle. Thus, a commonly used method is penalized quasi-likelihood (PQL), which is  implemented in packages such as lme4, nlme, and MASS. However, PQL relies on approximations of unknown accuracy to approximate the likelihood and suffers from problematic inferential properties, such as parameter estimates that tend to be too low (\ncite{mccu:sear:2001}). Since the likelihood is approximated to an unknown accuracy by the quasi-likelihood, any inference performed on the approximated likelihood will also produce results with an unknown level of accuracy.  Without bootstrapping, the user cannot know how valid their confidence intervals or likelihood ratio test results are when using PQL.  The popularity of PQL despite its inadequacies shows that there is a high demand for tools to fit GLMMs.

@@

My goals are  (1) to rewrite sections of my package in C to  improve its speed, (2) write functions to perform likelihood ratio tests for comparing nested models, (3) write additional functions to fit models with correlated random effects.

@@

(2)  Hypothesis testing for nested models can be split into three cases: the nested models differ in their fixed effects but have the same variance components, the nested models differ by one variance component and possibly some fixed effects, the nested models differ by two or more variance components and possibly by some fixed effects. I have worked out the details for calculating the test statistics and p-values for the first two cases and will  be able to write the code for these two in about three weeks.  Coding the last case will take longer because I will need to determine the test statistic and its sampling distribution.

@@

In conclusion, researchers from all fields have created a high demand for a tool to fit and test GLMMs. I would like to make the process of fitting a GLMM as user-friendly and theoretically-grounded as fitting a linear model. My package will allow researchers to find maximum likelihood estimates for GLMMs with various covariance structures and perform likelihood ratio tests to compare the fit of nested models.  This proposed work is an extension of my current  research on the topic of using MCLA to fit GLMMs.  I am the best individual for this project because I am familiar with MCLA's current limitations and areas of possible improvement.
